---
title: "06-The 'Ground Game'"
author: "Kento Yamada"
date: '2022-10-19'
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
knitr::opts_knit$set(root.dir = '/Users/kentoyamada/Documents/R-Projects/election-analytics-2022-midterms')
```

```{r pull functions, include = FALSE}
setwd("R")
files.sources = list.files()
sapply(files.sources, source)
setwd("..")
```

```{r load packages, include = FALSE}
library(tidyverse)
library(readxl)
library(sf)
library(stringi)
library(kableExtra)
library(stargazer)
library(usdata)
library(geofacet)
```

```{r load data, include = FALSE}
# Vote share
share_df <- read_csv("analysis_data/house_popvote_seats.csv") %>%
  select(-c("...1", "AreaAll")) # de-select unnecessary columns

# Expert ratings
ratings <- read_csv("analysis_data/expert_rating.csv")
# Expert ratings data for 2022
ratings_2022 <- read_excel("analysis_data/expert_ratings_2022.xlsx")

# Vote share by district
vote_share <- read_csv("analysis_data/house_party_vote_share_by_district_1948-2020.csv")

# Incumbency
incumb <- read_csv("analysis_data/incumb_dist_1948-2020.csv")
# Data for 2022
incumb_2022 <- read_csv("analysis_data/house_cands.csv")

# CVAP
cvap_by_2022_cd <- read_excel("analysis_data/cvap_by_2022_cd.xlsx")
```

```{r outcome by CD, include = FALSE}
d_win_df <- vote_share %>%
  # Filter out data from 2012 to 2018
  filter(raceYear >= 2012, raceYear <= 2018) %>% # Filter out 2018 data
  rename(year = raceYear, District = CD) %>% # Rename columns
  select(State, district_num, year, WinnerParty, District) %>%
  # 1 if Democrat won
  mutate(d_win = if_else(WinnerParty == "D", 1, 0))

# Find duplicates: summary(as.factor(d_win_df$District))
# Remove other duplicates caused by runoffs etc.
drop <- c(
  which(d_win_df$District == "LA-03" & d_win_df$year == 2016)[2],
  which(d_win_df$District == "LA-03" & d_win_df$year == 2012)[2],
  which(d_win_df$District == "LA-04" & d_win_df$year == 2016 & 
        d_win_df$WinnerParty == "D"),
  which(d_win_df$District == "LA-05" & d_win_df$year == 2014 & 
        d_win_df$WinnerParty == "D"),
   which(d_win_df$District == "ME-02" & d_win_df$year == 2018 & 
        d_win_df$WinnerParty == "R") # Ranked choice voting in ME
)
d_win_df <- d_win_df[-drop,]

# Change district names
d_win_df$District[which(d_win_df$District == "AK-AL")] <- "AK-01"
d_win_df$District[which(d_win_df$District == "DE-AL")] <- "DE-01"
d_win_df$District[which(d_win_df$District == "MT-AL")] <- "MT-01"
d_win_df$District[which(d_win_df$District == "ND-AL")] <- "ND-01"
d_win_df$District[which(d_win_df$District == "SD-AL")] <- "SD-01"
d_win_df$District[which(d_win_df$District == "VT-AL")] <- "VT-01"
d_win_df$District[which(d_win_df$District == "WY-AL")] <- "WY-01"

# Select relevant columns
d_win_df <- select(d_win_df, -c(State, district_num, WinnerParty))
```

```{r Democratic popular vote share 2012-2018, include = FALSE}
pop_share_df <- share_df %>%
  filter(year >= 2012, year <= 2018) %>%
  select(year, D_majorvote_pct) %>%
  rename(d_pop_share = D_majorvote_pct)
```

```{r clean expert ratings data, include = FALSE}
ratings_df <- ratings %>%
  # Convert state name to state abbreviation
  mutate(State = state2abbr(state)) %>%
  # District numbers
  mutate(District = str_pad(district, 2, pad = "0")) %>%
  # Define district names
  mutate(District = paste(State, District, sep = "-")) %>%
  
  # Filter out data from 2012 to 2018
  filter(year >= 2012, year <= 2018) %>%
  
  select(District, year, cook, rothenberg, sabatos_crystal_ball)

# Calculate ratings average
ratings_df$rating_avg = rowMeans(ratings_df[, -c(1,2)])

# Change district names
ratings_df$District[which(ratings_df$District == "AK-AL")] <- "AK-01"
ratings_df$District[which(ratings_df$District == "MT-AL")] <- "MT-01"
ratings_df$District[which(ratings_df$District == "ND-AL")] <- "ND-01"
ratings_df$District[which(ratings_df$District == "SD-AL")] <- "SD-01"
```

```{r further clean expert ratings data, include = FALSE}
# First merge d_win_df and ratings_df
ratings_adj_df <- left_join(d_win_df, ratings_df, by = c("District", "year"))

# If rating was missing for a particular district, that means that the race was not close
# Assume that the rating was 7 if the Republicans ended up winning and 1 if the Democrats ended up winning
ratings_adj_df[is.na(ratings_adj_df$rating_avg),]$rating_avg <- 
  if_else(ratings_adj_df[is.na(ratings_adj_df$rating_avg),]$d_win == 0, 7, 1)

# Select relevant columns
ratings_adj_df <- ratings_adj_df %>% select(year, District, rating_avg)
```

```{r load ads data, include = FALSE}
# ads <- read_csv("analysis_data/ads_2006_2018.csv")
# Compress and save object
# write_rds(ads, "analysis_data/ads_2006_2018.rds", compress = "xz")
ads <- readRDS("analysis_data/ads_2006_2018.rds")
ads_2022 <- read_excel("analysis_data/ads_2022_Sep19_Oct2.xlsx")
```

```{r clean ads historical data, include = FALSE}
ads_df <- ads %>%
  # De-select empty column
  dplyr::select(-1) %>% 
  # Select 2012-2018 elections
  filter(cycle >= 2012 & cycle <= 2018) %>% 
  rename(year = cycle) %>%
  mutate(District = substr(creative, start = 7, stop = 10)) %>%
  # Filter out non-Democrat/Republican ad
  drop_na(party) %>% 
  group_by(District, party, year) %>%
  # Number of ads per district & per party
  summarize(num_ad = n()) %>%
  # Pivot wider
  pivot_wider(names_from = party, values_from = num_ad,
              # Assume that there was 1 ad in districts where there were no ads for one party
              # Or else the log of the ratio will be infinity
              values_fill = 1) %>%
  # Log of ratio
  mutate(d_ad_log_ratio = log(Democrat/Republican)) %>%
  dplyr::select(year, District, d_ad_log_ratio, Democrat, Republican)

# Format district names
stri_sub(ads_df$District, 3, 2) <- "-"

# Select relevant columns and rows
ads_df <- select(ads_df, -c(Democrat, Republican))
ads_df <- ads_df[-which(ads_df$District == "US-01"),]
# IL-19 was eliminated due to redistricting
ads_df <- ads_df[-which(ads_df$District == "IL-19"),]
```

```{r clean incumbency data, include = FALSE}
# Filter out 2012-2018
incumb_df <-  filter(incumb, year >= 2012 & year <= 2018)

# Find duplicates: summary(as.factor(incumb_df$district_id))
# Remove other duplicates caused by runoffs etc.
drop_2 <- c(
  which(incumb_df$district_id == "LA03" & incumb_df$year == 2016)[2],
  which(incumb_df$district_id == "LA03" & incumb_df$year == 2012)[2],
  which(incumb_df$district_id == "LA04" & incumb_df$year == 2016)[1],
  which(incumb_df$district_id == "LA05" & incumb_df$year == 2014)[2],
  which(incumb_df$district_id == "ME02" & incumb_df$year == 2018)[2]
)
incumb_df <- incumb_df[-drop_2,]

# Code incumbent = 1
incumb_df <- incumb_df %>%
  mutate(d_inc = case_when(DemStatus == "Incumbent" ~ 1, 
                           TRUE ~ 0)) %>%
  rename(District = district_id) %>%
  dplyr::select(year, District, d_inc)
  
# Format district names
stri_sub(incumb_df$District, 3, 2) <- "-"

# Change district names
incumb_df$District[which(incumb_df$District == "AK-00")] <- "AK-01"
incumb_df$District[which(incumb_df$District == "DE-00")] <- "DE-01"
incumb_df$District[which(incumb_df$District == "MT-00")] <- "MT-01"
incumb_df$District[which(incumb_df$District == "ND-00")] <- "ND-01"
incumb_df$District[which(incumb_df$District == "SD-00")] <- "SD-01"
incumb_df$District[which(incumb_df$District == "VT-00")] <- "VT-01"
incumb_df$District[which(incumb_df$District == "WY-00")] <- "WY-01"
```


```{r join data sets, include = FALSE}
df <- full_join(d_win_df, ratings_adj_df, by = c("year", "District")) %>%
  full_join(ads_df, by = c("year", "District")) %>%
  full_join(incumb_df, by = c("year", "District"))

# If there were no ads in a district, we can say that there was no Democratic advantage
# in terms of ads. Thus, d_ad_log_ratio is coded =0.
df$d_ad_log_ratio[is.na(df$d_ad_log_ratio)] <- 0

# Democratic popular vote share in each year
df <- df %>%
  mutate(d_pop_share =
      case_when(year == 2012 ~ pop_share_df$d_pop_share[pop_share_df$year == 2012],
            year == 2014 ~ pop_share_df$d_pop_share[pop_share_df$year == 2014],
            year == 2016 ~ pop_share_df$d_pop_share[pop_share_df$year == 2016],
            year == 2018 ~ pop_share_df$d_pop_share[pop_share_df$year == 2018])
           )
```


```{r district-level model, include = FALSE}
model <- glm(d_win ~ d_pop_share+ rating_avg + d_ad_log_ratio + d_inc, 
             data = df, family = binomial)
```

```{r McFadden’s R-Squared, include = FALSE}
with(summary(model), 1 - deviance/null.deviance)
```

## Predicted Number of Democratic Seats

The data on the volume of TV ads was retrieved from the data that was compiled by Kantar/CMAG and analyzed by the Wesleyan Media Project (2022). This dataset shows the volume of TV ads aired between September 19 and October 2 in 53 House districts.

Ballotpedia (2022)


```{r clean 2022 expert ratings data, include = FALSE}
ratings_2022_df <- ratings_2022 %>%
  mutate(Cook = case_when(
    Cook == "Solid Democratic" ~ 1,
    Cook == "Likely Democratic" ~ 2,
    Cook == "Lean Democratic" ~ 3,
    Cook == "Toss-up" ~ 4,
    Cook == "Lean Republican" ~ 5,
    Cook == "Likely Republican" ~ 6,
    Cook == "Solid Republican" ~ 7
  )) %>%
  mutate(Inside = case_when(
    Inside == "Solid Democratic" ~ 1,
    Inside == "Likely Democratic" ~ 2,
    Inside == "Lean Democratic" ~ 3,
    Inside == "Tilt Democratic" ~ 3.5,
    Inside == "Toss-up" ~ 4,
    Inside == "Tilt Republican" ~ 4.5,
    Inside == "Lean Republican" ~ 5,
    Inside == "Likely Republican" ~ 6,
    Inside == "Solid Republican" ~ 7
  )) %>%
  mutate(Sabato = case_when(
    Sabato == "Safe Democratic" ~ 1,
    Sabato == "Likely Democratic" ~ 2,
    Sabato == "Lean Democratic" ~ 3,
    Sabato == "Toss-up" ~ 4,
    Sabato == "Lean Republican" ~ 5,
    Sabato == "Likely Republican" ~ 6,
    Sabato == "Safe Republican" ~ 7
  ))

# Calculate ratings average
ratings_2022_df$rating_avg = rowMeans(ratings_2022_df[, -1])
ratings_2022_df <- ratings_2022_df %>%
  dplyr::select(District, rating_avg)
```

```{r clean ads 2022 data, include = FALSE}
ads_2022_df <- ads_2022 %>%
  # Add "0" to district numbers
  mutate(District = str_pad(District, 2, pad = "0")) %>%
  # Change district names
  mutate(District = paste(State, District, sep = "-")) %>%
  
  # Assume that there was 1 ad in districts where there were no ads for one party
              # Or else the log of the ratio will be infinity
  mutate(`Pro-Rep Airings` = 
           if_else(`Pro-Rep Airings` == 0, 1, `Pro-Rep Airings`)) %>%
  mutate(`Pro-Dem Airings` = 
           if_else(`Pro-Dem Airings` == 0, 1, `Pro-Dem Airings`)) %>%
  
  # Log of ratio
  mutate(d_ad_log_ratio = log(`Pro-Dem Airings`/`Pro-Rep Airings`)) %>%
  dplyr::select(District, d_ad_log_ratio)
```


```{r clean incumbency data 2022, include = FALSE}
incumb_2022_df <- incumb_2022 %>%
  # Convert state name to state abbreviation
  mutate(State = state2abbr(state)) %>%
  # District numbers
  mutate(District = str_pad(district, 2, pad = "0")) %>%
  # Define district names
  mutate(District = paste(State, District, sep = "-"))

# Change district names
incumb_2022_df$District[which(incumb_2022_df$District == "AK-AL")] <- "AK-01"
incumb_2022_df$District[which(incumb_2022_df$District == "DE-AL")] <- "DE-01"
incumb_2022_df$District[which(incumb_2022_df$District == "WY-AL")] <- "WY-01"
incumb_2022_df$District[which(incumb_2022_df$District == "VT-AL")] <- "VT-01"
incumb_2022_df$District[which(incumb_2022_df$District == "ND-AL")] <- "SD-01"
incumb_2022_df$District[which(incumb_2022_df$District == "SD-AL")] <- "SD-01"

# Democratic incumbent running for re-election
incumb_2022_df <- incumb_2022_df %>%
  mutate(d_inc = case_when(cand_party == "Democratic" & incumbent == 1 ~ 1, 
                           TRUE ~ 0)) %>%
  filter(d_inc == 1) %>%
  # Select relevant columns
  select(District, d_inc) 
```


```{r merge 2022 data, include = FALSE}
df_2022 <- ratings_2022_df %>%
  full_join(ads_2022_df, by = "District") %>%
  full_join(incumb_2022_df, by = "District")

# If there were no ads in a particular district, the Democratic advantage is 0
df_2022$d_ad_log_ratio[is.na(df_2022$d_ad_log_ratio)] <- 0

# d_inc = 0 if no Democratic incumbent
df_2022$d_inc[is.na(df_2022$d_inc)] <- 0

# Predicted popular vote share
df_2022$d_pop_share <- 47.902 # This is the prediction from Blog Post 5
```

```{r district-level prediction for 2022, include = FALSE}
df_2022$pred <- predict(model, df_2022, type = "response")
```

```{r clean cvap data, include = FALSE}
colnames(cvap_by_2022_cd) <- c("District", "CVAP")
```

```{r simulated distribution of draws from CVAP, include = FALSE}
# Number of draws
draw <- 10000

# List of districts
district_list <- df_2022$District

# Empty matrix
sim_D_votes <- matrix(NA, nrow = 435, ncol = draw + 2)
# First column: Names of districts
sim_D_votes[,1] <- df_2022$District

set.seed(12345)
# Get predicted distribution from CVAP in each district
for(i in 1:length(district_list)){
  # Second column: CVAP
  sim_D_votes[i,2] <-cvap_by_2022_cd$CVAP[cvap_by_2022_cd$District == district_list[i]]
  
  # Third column and beyond: predicted probability
  sim_D_votes[i, 3:(draw+2)] <- 
  rbinom(n = draw,
         # CVAP of District i
         size = cvap_by_2022_cd$CVAP[cvap_by_2022_cd$District == district_list[i]], 
         # Predicted probability for D in District i
         prob = df_2022$pred[df_2022$District == district_list[i]])
}
```

```{r CA 22 distribution, echo = FALSE, message = FALSE, warning = FALSE}
# Filter out predictions for CA-22
sim_D_votes[df_2022$District == "CA-22",] %>%
  tail(draw) %>% # Draw predictions
  as.numeric() %>%
  as.tibble() %>%
  ggplot(aes(x = value)) +
  geom_histogram(binwidth = 100) +
  labs(title = "CA-22: Predicted Number of Votes for Democratic Candidate",
       subtitle = "10,000 Binomial Process Simulations",
       x = NULL, y = "Frequency")
```

```{r prepare data frame for plotting, include = FALSE}
# Filter out predictions for Arizona
sim_D_votes_az <- sim_D_votes[grep("AZ", df_2022$District),] %>%
  as_tibble()

# Pivot longer
colnames(sim_D_votes_az)[1] <- "District"
colnames(sim_D_votes_az)[2] <- "CVAP"
sim_D_votes_az <- sim_D_votes_az %>%
  pivot_longer(cols = 3:(draw + 2)) %>%
  mutate(value = as.numeric(value))
```

```{r make grid, include = FALSE, message = FALSE, warning = FALSE}
az_grid <- data.frame(
  row = c(1, 1, 2, 2, 3, 3, 3, 4, 4),
  col = c(1, 3, 1, 2, 1, 2, 3, 1, 3),
  name = (rep("", 9)),
  code = c("AZ-09", "AZ-02", "AZ-08", "AZ-01", "AZ-03", "AZ-04", "AZ-05", "AZ-07", "AZ-06"),
  stringsAsFactors = FALSE
)
```

```{r faceted histogram, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(sim_D_votes_az, aes(x = log(value))) +
  geom_histogram(bins = 500) +
  facet_geo(~ District, grid = az_grid) +
  labs(title = "Predicted Votes Cast for Democratic Candidates",
        x = "Natural Log of Democratic Votes")
```

```{r predicted probability, echo = FALSE}
ggplot(df_2022, aes(x = pred)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(aes(xintercept = 0.5), color = "purple") +
  theme_bw() +
  labs(title = "Predicted Probability of Democratic Victory in Each District",
       x = "Predicted Probability of Democratic Victory",
       y = "Frequency")
```

```{r exclude safe seats, echo = FALSE}
df_2022 %>%
  filter(rating_avg != 7 & rating_avg != 1) %>%
  ggplot(aes(x = pred)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(aes(xintercept = 0.5), color = "purple") +
  theme_bw() +
  labs(title = "Predicted Probability of Democratic Victory in Each District",
       subtitle = "Excluding Safe Seats",
       x = "Predicted Probability of Democratic Victory",
       y = "Frequency")
```

```{r count number of predicted Democratic victory, include = FALSE}
sum(df_2022$pred > 0.5)
```

```{r count number of Republican victory, include = FALSE}
435 - sum(df_2022$pred > 0.5)
```


### References

Bafumi, J., Erikson, R., & Wlezien, C. (2018). Forecasting the 2018 Midterm Election using National Polls and District Information. *PS: Political Science & Politics, 51*(S1), 7-11. doi:10.1017/S1049096518001579

Ballotpedia. (2018). United States House of Representatives elections, 2018. https://ballotpedia.org/United_States_House_of_Representatives_elections,_2018

Ballotpedia. (2022). United States Congress elections, 2022. https://ballotpedia.org/United_States_Congress_elections,_2022

Cook Political Report. (2022). PVI Map and District List. https://www.cookpolitical.com/cook-pvi/2022-partisan-voting-index/district-map-and-list

Gerber, A.S., Gimpel, J. G., Green, D. P., & Shaw, D. R. (2011). How Large and Long-lasting Are the Persuasive Effects of Televised Campaign Ads? Results from a Randomized Field Experiment. *American Political Science Review, 105*(1), 135–150. https://doi.org/10.1017/S000305541000047X

Huber, G.A. & Arceneaux, K. (2007). Identifying the Persuasive Effects of Presidential Advertising. *American Journal of Political Science, 51*(4), 957–977. https://doi.org/10.1111/j.1540-5907.2007.00291.x

Wesleyan Media Project. (2022, October 6). Democrats Out-Pacing GOP in Senate Races. https://mediaproject.wesleyan.edu/releases-100622/

